### 2020-07-02 chapters/sota.tex
Restivo feedback applied

A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}. In this Chapter, we discuss the most relevant work for the problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems, presented in Chapter \ref{chap:intro}. We present a wide analysis of several categories of algorithms that seem viable to build the desired system.

%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought to solve the problem and the sections

\section{Outlier detection in time-series data} \label{sec:outliers}
Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.

@@

Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy} \cite{Blazquez-Garcia-Review-Anomaly-Detection}) .

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=1]{figures/taxonomy-point-outlier-uni.png}
    \caption{Point outlier detection in univariate time-series taxonomy}
    \label{fig:point-outlier-taxonomy}
  \end{center}
\end{figure}

@@

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. For example, nearest neighbor based methods consider a datum an outlier if it has less than $\alpha$ neighbors. Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

@@

\subsubsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm} \label{alg:extran}

Yang et. al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios. The authors propose \emph{Extra-N}, an algorithm that detects point outliers based on the number of neighbors it has. The authors make use of the \emph{predictability} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.

The authors of Extra-N want to maintain for each data point $p\textsubscript{i}$ the number of neighbors it has, rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors. The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers. The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors. The authors introduce the concept of \textit{lifetime neighbor counts} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where it exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window. Consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors: \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, 
\[ p\textsubscript{i}.lt\_cnt = \{W\textsubscript{i}: 3, W\textsubscript{i+1}: 2, W\textsubscript{i+2}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.

@@

\subsubsection*{Space complexity}
Extra-N has to keep a lifetime neighbor count dictionary-like object for each window data point. For that reason, we claim that Extra-N achieves linear memory consumption relative to the number of data points in the window.

\subsubsection*{Applicability to our Hypothesis}
Extra-N memory consumption grows linearly in regards to window size, so it does not fit our desired lightweight system architecture. Furthermore, updating the sliding window should be done in constant time, which is not the case for Extra-N. Hence, this algorithm will be of no use to us.

@@

HTM is an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream. In other words, if there is a sudden abrupt change in the distribution of a specific time-dependent variable, the algorithm will raise an alert. However, if that initially thought to be random spike becomes frequent, the algorithm adapts and assimilates this as "a new reality" and stops classifying it as an anomaly. Unfortunately, this property is not desirable in our system. Our stream monitoring system must report changes relatively to an initial static configuration (reference or normality period) and should not accept a new distribution as the new standard.

@@

DeepAnT employs a Convolutional Neural Network (CNN) as its time-series predictor module. Before monitoring a data stream, the CNN model must be trained. The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN. Additionally, the data set used to train the model thus not require labels.

@@

\subsubsection*{Applicability to our Hypothesis}
Similar to the HTM algorithm presented, DeepAnT adapts to the underlying distribution of the data stream. In other words, if the data patterns shift from the initial configuration but stabilize, DeepAnT will accept it as it is. As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again. Furthermore, DeepAnT uses deep learning which is actually a subset of machine learning and just as inexplicable. In summary, we do not deem DeepAnT usable for our use case.

@@

\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}

@@

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of-magnitude faster than some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection{Membership Queries and the Bloom filter} \label{sec:bloom}
% approx aggregation to compute
A Bloom filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1.1]{figures/initial-bloom.png}
      \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/insert-bloom.png}
      \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
      \label{fig:insertion-bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/query-bloom.png}
      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

@@

\subsection{Item Frequency and the Count-Min Sketch}
% approx aggregation to compute
The Count-Min Sketch (CMS) uses two principles from the discussed Bloom filters in Section \ref{sec:bloom}. First, Bloom filters show that precision can be sacrificed to achieve space savings. Second, they are very similar at a technical level.

@@

\subsection{Cardinality Estimation and the HyperLogLog} \label{sec:hyper-log-log}

@@

\section{Sliding Window Aggregations with Probabilistic Data Structures} \label{sec:sliding-pds}
In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space. These structures aggregate an endless stream of data. However, when working with sliding windows there is the need not only to insert new events but to evict old ones. The probabilistic data structures described until now can aggregate new incoming data but do not address the type of temporality needed for sliding windows --- \textit{i.e.} the ability to expire old elements from the aggregated state.

@@

>>>>

A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams, Domingos-Mining-Time-Data-Streams}. In this chapter, we discuss the most relevant work for the problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems, presented in Chapter \ref{chap:intro}. We present a wide analysis of several categories of algorithms to be able to identify what's the region of this vast landscape of methods that may be useful to build the desired system.

\section{Outlier detection in time-series data} \label{sec:outliers}
Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey, Aguinis-Outlier-survey, Chandola-Outlier-survey-2009, Hodge-Outlier-survey, Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.

@@

Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy} \cite{Blazquez-Garcia-Review-Anomaly-Detection}) .
\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=1]{figures/taxonomy-point-outlier-uni.png}
    \caption{Point outlier detection in univariate time-series taxonomy}
    \label{fig:point-outlier-taxonomy}
  \end{center}
\end{figure}

@@

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. For example, nearest neighbor based methods consider a datum an outlier if it has less than $k$ neighbors. Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

@@

\subsubsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data} \label{alg:extran}

Yang et al. propose \emph{Extra-N} \cite{Yang-Neighbor-Based-Pattern-Detection}, a method for incremental detection of point outliers in sliding window scenarios. Data points are considered outliers if they have few neighbors, \textit{i.e.}, less than a user-defined threshold $\alpha$ of neighbors within a user-defined range $\theta$. Figure \ref{fig:point-outlier} depicts points plotted in their feature space (in this case only two dimensions or features are used, x1 and x2) and a point outlier for $alpha=2$ points and any given $\theta$.
\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=0.6]{figures/point-outlier.pdf}
    \caption{Point outlier for $\alpha=2$ and a given $\theta$}
    \label{fig:point-outlier}
  \end{center}
\end{figure}

The key idea is to predict the expiration of any data point in the sliding window and pre-handle the impact it has on its neighbors. The authors introduce the concept of \emph{lifetime neighbor counts} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point corresponds to a sequence of predicted neighbor counts, \textit{i.e.}, the number of predicted neighbors a data point has in any future window where it exists. More precisely, each entry on \textit{lt\_cnt} records the number of neighbors that are known to survive in future windows.

For example, consider a window \textit{W\textsubscript{0}} and a data point \textit{p\textsubscript{0}} with three neighbors: \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{0}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{1}} and \textit{p\textsubscript{0}} expires after \textit{W\textsubscript{2}}. This implies that at \textit{W\textsubscript{0}}, 
\[ p\textsubscript{0}.lt\_cnt = \{W\textsubscript{0}: 3, W\textsubscript{1}: 2, W\textsubscript{2}: 0\} \]
The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point check if the first or current element of \textit{lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether that data point is an outlier or not.

@@

\subsubsection*{Space complexity}
Extra-N has to keep a lifetime neighbor count dictionary-like object for each window data point. For that reason, Extra-N achieves linear memory consumption relative to the number of data points in the window.

\subsubsection*{Applicability to our Hypothesis}
Extra-N memory consumption grows linearly in regards to window size, so it does not fit our desired lightweight system architecture. Furthermore, updating the sliding window should be done in constant time, which is not the case for Extra-N. Hence, we will not use this algorithm.

@@

HTM is an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream. In other words, if there is a sudden abrupt change in the distribution of a specific time-dependent variable, the algorithm will raise an alert. However, if an initial spike that was assumed to be an anomaly becomes frequent, then the algorithm adapts and assimilates this as "a new reality" and stops classifying it as anomalous. Unfortunately, this property is not desirable in our system. Our stream monitoring system must report changes relatively to an initial static configuration (reference or normality period) and should not accept a new distribution as the new standard.

@@

DeepAnT employs a Convolutional Neural Network (CNN) as its time-series predictor module. Before monitoring a data stream, the CNN model must be trained. The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN. Additionally, the data set used to train the model does not require labels.

@@

\subsubsection*{Applicability to our Hypothesis}
Similar to the HTM algorithm presented, DeepAnT adapts to the underlying distribution of the data stream. In other words, if the data patterns shift from the initial configuration but stabilize, DeepAnT will accept it as it is. As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again.

@@

\section{Sliding window aggregation algorithms} \label{sec:sota-swag-algs}

@@

\section{Probabilistic data structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of-magnitude faster than some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection{Membership queries and the Bloom Filter} \label{sec:bloom}
% approx aggregation to compute
A Bloom Filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.9]{figures/initial-bloom.png}
      \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/insert-bloom.png}
      \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
      \label{fig:insertion-bloom-filter}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/query-bloom.png}
      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

@@

\subsection{Item frequency and the Count-Min Sketch}
% approx aggregation to compute
The Count-Min Sketch (CMS) uses two principles from the discussed Bloom filters in Section \ref{sec:bloom}. First, Bloom filters show that precision can be sacrificed to achieve space savings. Second, they are very similar at a technical level.

@@

\subsection{Cardinality estimation and the HyperLogLog} \label{sec:hyper-log-log}

@@

\section{Sliding window aggregations with probabilistic data structures} \label{sec:sliding-pds}
In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space. These structures aggregate an endless stream of data. However, when working with sliding windows there is the need not only to insert new events but to evict old ones. The probabilistic data structures described until now can aggregate new incoming data but do not address the type of temporality needed for sliding windows --- \textit{i.e.} the ability to expire old elements from the aggregated state.

@@

\section{Summary}
